---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

demrat
=======

Provides a framework for estimating demographic rates (growth, crude birth, and total fertility rate) of a population represented by a skeletal sample(s) using age-at-death ratios. 

The methodology accounts for the effect of stochastic variation in small skeletal samples and allows to set a reference mortality pattern appropriate for the time period being studied.


Shiny web application
------------------------

[Shiny app for the `demrat` package](https://galetap.shinyapps.io/demrat/) enables to deploy demographic predictions using `demrat` with a point-and-click interface. The Shiny app is useful for non-R users or for those who want to experiment with the `demrat` package before downloading it and without having to read through the documentation below.

Installation and loading
------------------------

-   Install the latest developmental version from
    [GitHub](https://github.com/galetap/demrat) as follow:

```{r installation, eval = FALSE}
if(!require(devtools)) install.packages("devtools")
devtools::install_github("galetap/demrat")
```

-   Loading package(s)

```{r loading_package}
library(demrat)
# The examples below use the `tidyverse` package. If installed, load it by running:
library(tidyverse)
```


Data
------------------------

The functions of the package accept data in two formats; summary data and raw data format. 

Two built-in datasets are available: `BA` (summary data format) and `BAraw` (raw data format). Both datasets are adopted from Bocquet-Appel (2002), Table 1 (https://doi.org/10.1086/342429). 


**Summary data format**

The rows of the summary data format contain data for one skeletal sample. Data must have at least four variables. 

1. Site
2. Culture
3. D20_ (i.e., D20+, number of adult skeletons in a sample)
4. One out of the following three age-at-death ratios: D1_D20_, D3_D20_, D5_D20_ (i.e., D1+/D20+, D3+/D20+, D5+/D20+ ratios, respectively).

The column with the total number of skeletons in a sample (named n) is recommended. 

```{r Summary_format}
tibble(Site = c("Site A", "Site B", "Site C"),
       Culture = c("Culture 1", "Culture 1", "Culture 2"),
       D20_ = c(25, 96, 55),
       D5_D20_ = c(1.61, 1.37, 1.27),
       n = c(50, 141, 72))
```

*`BA` dataset*

Summary format of Bocquet-Appel's (2002) dataset. 

Each row contains data from a single skeletal sample. 

Variables include Site and Culture name, absolute chronology of the Site, number of skeletons in various age-at-death categories, and age-at-death ratio values.   

```{r Dataset_BA}
BA
# For column description, run help(BA).
```


**Raw data format**

The rows of the raw data format contain data for one skeleton. Data must contain at least four variables. 

1. Site
2. Culture
3. Age_min; lower limit of the age-at-death estimate of a skeleton
4. Age_max; upper limit of the age-at-death estimate of a skeleton

Note that the age-at.death intervals are closed on the left and open on the right.
It is, interval from 20 to 40 means that individual died between 20 and 39.9 years of age.  

```{r Raw_format}
# First 10 rows only
tibble(Site = rep("Site A", 10),
       Culture = rep("Culture 1", 10),
       Age_min = c(20, 0, 40, 3, 9, 20, 1, 20, 19, 60),
       Age_max = c(40, 1, 60, 5, 12, 39, 2, 100, 21, 100))
```

*`BAraw` dataset*


Raw format of Bocquet-Appel's (2002) dataset. 

```{r Dataset_BAraw}
BAraw
# For column description, run help(BAraw).
```


Key functions
-------------

#### `dr`: Translate raw to summary data format

Based on the raw data format, the function calculates 

* number of skeletons in several age-at-death groups (e.g., D1+, D20+) and 
* six demographic ratios (D1+/D20+, D3+/D20+, D5+/D20+, D5-14/D20+ aka juvenility index, D5-19/D5+ aka P index, and D0-14/D0+) by Site and Culture.

The function can be applied for one or more Sites. 

```{r dr}
dr(BAraw)
```


#### `diest`: Estimates demographic rates for skeletal sample(s)

`diest` function estimates growth, crude birth, and total fertility rates based on age-at-death ratios (e.g., D5+/D20+) by Site and Culture. The function accepts data in the summary format. 

The function can be applied for one or more Sites. 

Estimation for sites with a large number of skeletons and/or estimation for many sites can be slow. 

**`diest` with default settings based on summary data format**

By default, the prediction model is based on the reference set of 100 simulated reference skeletal samples 

* that have the same number of adult skeletons (D20+) as the real skeletal sample under study
* that are drawn from reference populations with life expectancies between 18 and 25 years and subjected to an annual growth between −3 and 3% per annum.

```{r diest_summary_format}
BA %>%
  # Subset two sites only
  slice(12, 24) %>%
  # Predict demographic rates
  diest()
```

**`diest` with user defined settings of demographic regime of reference populations**

A demographoc regime (mortality level and annual growth) of a population from which reference skeletal samples are drawn can be set by the user (run `help(diest)`).

```{r diest_set}
BA %>%
  slice(12) %>%
  diest(samples = 200, e0_min = 25, e0_max = 30)
```


**`diest` with subsettings dependent and independent variables**

By setting the `IV` and `DV` arguments, a subset of results can be displayed.

```{r diest_filter}
BA %>%
  slice(12) %>%
  diest(IV="D5_D20_", DV = "TFR")
```


**`diest` with full prediction results**

Full results of the prediction can be obtained by setting `summary=F`.

```{r diest_full}
BA %>%
  slice(12) %>%
  diest(summary=F)
```


And further analyzed by `unnest()`.

```{r diest_full_unnest}
BA %>%
  slice(12) %>%
  diest(summary=F) %>% 
  unnest(DIest)
```


**`diest` based on raw data format**

Raw data format must be converted to summary data format before using `diest`.

```{r diest_raw_format}
BAraw %>%
  # Convert raw to summary data format (compute demographic ratios)
  dr() %>% 
  # Select data for two sites
  slice(12, 24) %>%
  # Predict demographic variables
  diest()
```


#### `plot_diest`: Plots prediction model and estimation for a selected site

The function visualizes the prediction model. The plot has an age-at-death ratio (independent variable) on the X axis and a demographic variable (dependent variable) on the Y axis. The scatter shows the reference set of simulated skeletal samples (points), regression line used for a prediction, and the age-at-death ratio and predicted variable for a given Site. 

The function accepts data in the summary format.

The age-at-death ratio (D5+/D20+ as default), estimated demographic variable (Growth as default), and demographic pattern of reference populations (see diest() for defaults) can be set by the user.  

The function can be applied for one Site only. 

**`plot_diest` with default settings**

```{r plot_diest_default}
BA %>%
slice(12) %>%
plot_diest()
```


**`plot_diest` with user-defined settings**

```{r plot_diest_user}
BA %>%
slice(12) %>%
plot_diest(DV = "TFR", samples=500, e0_min=25, e0_max=30)
```


#### `simdr_CD`: Creates a reference set of simulated skeletal samples

`simdr_CD` is a helper function that creates a reference set of skeletal samples drawn from Coale and Demeny (1983) set of model life tables. Each row contains one simulated skeletal sample. For parameters description, run `help(simdr_CD)`.

**`simdr_CD` with default settings**

```{r simdr_CD_default}
# Simulation based on the default settings 
simdr_CD() %>% 
  as_tibble()
```


**`simdr_CD` with user-defined settings**

```{r simdr_CD_user}
simdr_CD(samples = 200, D20_raw = 20, e0_min = 30, e0_max = 40) %>% 
  as_tibble()
```


**Using `simdr_CD` to study a relationship between demographic variables**

Since `simdr_CD` generates reference skeletal samples or populations with various life expectancies at birth and annual growth rates, it can be used to depict a relationship between demographic variables.

```{r simdr_CD_TFR_Growth, fig.width=5, fig.height=4}
simdr_CD() %>% 
  ggplot(aes(x = TFR, y = Growth)) + 
  geom_point() + 
  labs(title = "Relationship between TFR and annual growth rate(%)",
       subtitle = "Assuming a life expectancy at birth between 20 and 30 years\nand 50 adult deaths in each skeletal sample") + 
  scale_x_continuous(breaks = seq(0, 20, 1)) + 
  theme_classic() + 
  theme(axis.title = element_text(size = 8),
        plot.title = element_text(size = 10, face = "bold"),
        plot.subtitle = element_text(size = 9))
```

  

```{r}


```
